---
title: "cm008 Exercises"
output: github_document

---

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))

```

## `mutate()`

Let's get: 

- GDP by multiplying GPD per capita with population, and
- GDP in billions, named (`gdpBill`), rounded to two decimals.

```{r}
## mutate() is used to add a new column round() is round it to 2 decimal
temp <- gapminder %>% 
  mutate(gdp = gdpPercap * pop,
         gdpBill = round(gdp/10^9,2))
#knitr::kable(temp)

```


Notice the backwards compatibility! No need for loops!

Try the same thing, but with `transmute` (drops all other variables). 

```{r}
gapminder %>% 
  transmute(gdp = gdpPercap * pop,
         gdpBill = round(gdp/10^9,2))

```

The `if_else` function is useful for changing certain elements in a data frame.

Example: Suppose Canada's 1952 life expectancy was mistakenly entered as 68.8 in the data frame, but is actually 70. Fix it using `if_else` and `mutate`. 

```{r,results='hold'}
##hold all the output pieces and push them to the end of a chunk

if_else(letters == "h", "H", letters)
## this is replacing h with H
temp <- gapminder %>%
  mutate(lifeExp = if_else(country == "Canada" & year == 1952, 70, lifeExp)) %>% 
  filter(country == "Canada")
knitr::kable(temp)

```


Your turn: Make a new column called `cc` that pastes the country name followed by the continent, separated by a comma. (Hint: use the `paste` function with the `sep=", "` argument).

```{r}
temp<- gapminder %>% 
  mutate(cc=paste(country, continent,sep=","))
knitr::kable(temp)

```

These functions we've seen are called __vectorized functions__. 

## `summarize()` and `group_by()`

Use `summarize()` to compute the mean and median life expectancy using all entries:

```{r}
gapminder %>%
  summarize(avg_lifeExp = mean(lifeExp),
            median_lifeExp =median(lifeExp))
```

Do the same thing, but try:

1. grouping by country
2. grouping by continent and country

```{r}
gapminder %>%
  group_by(country) %>% 
  summarize(avg_lifeExp = mean(lifeExp),
            median_lifeExp =median(lifeExp))


gapminder %>%
  group_by(continent) %>% 
  summarize(avg_lifeExp = mean(lifeExp),
            median_lifeExp =median(lifeExp))

gapminder %>% 
  group_by(continent, country)

gapminder %>%
  group_by(country,continent) %>% 
  summarize(avg_lifeExp = mean(lifeExp),
            median_lifeExp =median(lifeExp))


```

- Notice the columns that are kept. 
- Notice the grouping listed above the tibble, especially without a call after grouping.
- Notice the peeling of groups for each summarize. 

Question: What if I wanted to keep the other numeric columns (gdpPercap, pop)? Can I? Would this even make sense?

For each continent: What is the smallest country-wide median GDP per capita?

```{r}

## smallest country-wide median GDP per capita
gapminder %>% 
  group_by(continent,country) %>% 
  summarize(md = median(gdpPercap))

## the smallest GDP per capita within each continent
gapminder %>% 
  group_by(continent,country) %>% 
  summarize(md = median(gdpPercap)) %>% 
  summarize(min=min(md))

```

Note that ggplot2's grouping is different from dplyr's! Try making a spaghetti plot of lifeExp over time for each coutry, by piping in a grouped data frame -- it won't work:

```{r}
gapminder %>% 
  group_by(country) %>% 
  ggplot(aes(lifeExp,year))+
  geom_line()
```


Your turn! For each continent, what is the median GDP per capita of countries with high (>60) life expectancy vs countries with low (<=60)? Sort this data frame by median GDP per capita.

```{r}
gapminder %>%
  group_by(country,lifeExp>60) %>% 
  summarize(md=median(gdpPercap)) %>% 
  arrange(md)

## insert an column to state the high and low level of lifeExp, when lifeExp>60 indicate high orelse indicate low
gapminder %>% 
  mutate(age = if_else(lifeExp > 60,"high","low")) %>% 
  group_by(continent, age) %>% 
  summarise(md = median(gdpPercap)) %>% 
  arrange(md)


```

There are special functions to summarize by. Let's see some of them:

- `n()`: Number of rows in the group.
- `n_distinct()`

Convenience functions:

- `tally()` (= `summarize(n = n())`)
- `count(...)` (= `group_by(...) %>% tally()`)

n_distinct: How many years of record does each country have?

```{r}
## count thw number of record in each continent
gapminder %>% 
  group_by(continent) %>% 
 # summarize(n=n())
  tally()

gapminder %>% 
  count(continent)



```

count

## Function types

Let's stop to identify some theory of function types, and the `dplyr` approach to them.

| Function type | Explanation | Examples | In `dplyr` |
|------|-----|----|----|
| Vectorized functions | These take a vector, and operate on each component independently to return a vector of the same length. In other words, they work element-wise. | `cos`, `sin`, `log`, `exp`, `round` | `mutate` |
| Aggregate functions | These take a vector, and return a vector of length 1 | `mean`, `sd`, `length`, `typeof` | `summarize`, esp with `group_by`. |
| Window Functions | these take a vector, and return a vector of the same length that depends on the vector as a whole. | `lag`, `rank`, `cumsum` | `mutate`, esp with `group_by` |

For any generic output, we can use dplyr's `do()` function -- but that's a topic for STAT 547. 

## Grouped `mutate()`

Calculate the growth in population since the first year on record _for each country_. `first()` is useful.

```{r}

gapminder %>% 
  group_by(country) %>% 
  mutate(growth = pop - first(pop))   ## compare it within the group

gapminder %>%
  group_by(country) %>% 
  mutate(change =pop - lag(pop))   ## do the same things  find the difference between each entry with the first one
```

Notice that `dplyr` has retained the original grouping.

How about growth compared to `1972`?

```{r}

gapminder %>% 
  group_by(country) %>% 
  mutate(change= pop-pop[year== 1972])

```

Make a new variable `pop_last_time`, as the "lag-1" population -- that is, the population from the previous entry of that country. Use the `lag` function.

```{r}
gapminder %>% 
  group_by(country) %>% 
  mutate(pop_last_time= pop-lag(pop,1))
```

Similar: `lead` function.

Notice the NA's. 

## Putting it all together

Your turn: Use what we learned to answer the following questions.

1. Determine the country that experienced the sharpest 5-year drop in life expectancy, in each continent.

```{r}
##filter out the entry which has no gain results using the !is.na(gain)
## 2 min() function is used to get the sharpest drop in each ontinent
gapminder %>% 
  group_by(continent,country) %>% 
  mutate(gain=lag(lifeExp)-lifeExp) %>% 
  filter(!is.na(gain)) %>% 
  summarize(min=min(gain)) %>% 
  summarize(min=min(min)) %>% 
  arrange(min)

```

2. Compute the relative gdp (NOT per capita!) of each country compared to Canada (= GDP of a country / GDP of Canada).

```{r}
## GDP of canada
cda <- gapminder %>% 
  mutate(gdp=gdpPercap* pop) %>% 
  filter(country=="Canada") %>% 
  `[[`("gdp")
#knitr::kable(cda)

## rep() replicate the value in the function
## nlevels return the levels the argument has
my_df <- gapminder %>% 
  mutate(cda_gdp=rep(cda, nlevels(country)),
         gdp=gdpPercap*pop,
         rel_gdp=gdp/cda_gdp)  ## replicate
knitr::kable(my_df)

                       
my_df %>%  filter(country == "Canada")
summary(my_df$rel_gdp)


```

Sanity check: are Canada's numbers = 1? What is the spread of numbers like (should be small)?

## Summary of major one-table functions

- `select()`
- `filter()`
- `arrange()`
- `mutate()`
- `summarize()`

Together with `group_by()` and "twists" of the above. 


## Practice Exercises

Practice these concepts in the following exercises. It might help you to first identify the type of function you are applying.

1. Convert the population to a number in billions.

2. Compute the change in population from 1962 to 1972 for each country.

3. Rank the continents by GDP per capita. You should have two columns: one with continent, and another with the ranking (1 through 5). __Hint__: use the `rank()` or `min_rank()` function.
